Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-08-04 paper/manuscript.tex
last changes

\begin{frontmatter}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Abstract
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist in as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest, permits the robust computation of free energy for large transformations.
Alchemical calculations can be used in a variety of scenarios, such as: 
\begin{itemize}
\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{A});
\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 
\end{itemize}

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
\end{itemize}

After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate the free energy for the transformation of interest.
 Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}. %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine} and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform. 
In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
This Best Practices guide provides current recommendations and tips for users of all experience.

@@

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
What can be expected from this Best Practices guide is: a good starting point for new practitioners and a reference for experienced practitioners. 
 For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently unsettled, we highlight areas where further research is needed to identify an unambiguous recommendation.
 It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

 What we assume from a novice practitioner is a moderate experience with molecular simulation concepts and use of simulation packages. 
 Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

 While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead the guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical calculation? 
     \item How do I select an appropriate alchemical protocol? 
     \item What software tools are available to perform alchemical calculations? 
     \item How should I analyze my data and report uncertainties? 
\end{itemize}

Some other background information may be needed depending on the nature of the alchemical project. For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations. This will require some basic familiarity on how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}.

@@

\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$L$ and $R$ --- generic names for ligand and receptor\\
$K_b^{\circ}$ --- binding  constant \\
$c^{\circ}$ --- standard state concentration \\
$U$ --- potential energy\\
$u$ --- reduced potential describing a thermodynamic state \\
$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble)\\
$\Delta A$ --- Helmholtz free energy (canonical ensemble)\\
$\Delta f$ --- reduced (dimensionless) free energy \\
$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy\\
$\Gamma$ --- conformation space accessible by simulations \\
$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
$k_B$ --- Boltzmann constant \\
$Z$ --- partition function \\
$p$ --- pressure \\
$\mu$ --- chemical potential (grand canonical ensemble)\\
$T$ --- temperature \\
$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
$\vec{\lambda}$ --- alchemical progress parameter \\
$g$ --- statistical inefficiency\\
$\mathcal{O}$ --- overlap matrix\\
$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
$\tau _{eq}$ --- integrated auto-correlation time\\
$t_0$ --- equilibration time
\end{tcolorbox}

%%%%%%%%%%%%%%%%%%%%
% Theory basics    %
%%%%%%%%%%%%%%%%%%%%
\section{Statistical mechanics demonstrates why alchemical free energy calculations work}
\label{sec:theory}
Why would you want to run an alchemical free energy calculation and why do they work?
In this section, we use the example of relative free energy calculation to sketch the theory of alchemical calculations and illustrate their utility.
The emphasis here is placed on bridging theoretical foundations and intuition.
A rigorous derivation of the standard (absolute) free energy of binding using the principles of statistical mechanics can be found in Gilson's classic work~\cite{gilson1997statisticalthermodynamic}.

@@

\paragraph{The free energy of binding can be expressed as a ratio of partition functions}
A natural though generally very computational expensive way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
\begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
    \Delta G_{\mathrm{bind},L} \approx \Delta A_{\mathrm{bind}, L} = - k_B T  \left( \mathrm{ln} \frac{P(RL)}{P(R+L)} + \ln \left( c^\circ N_{\mathrm{Av}} V \right) \right) \, ,
\end{equation}
where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand conformations $\vec{q}$ that we consider bound and unbound respectively.
The probability of a conformation $\vec{q}$ is given by the Boltzmann probability density function
\begin{equation}
    P(\vec{q}) = \frac{\exp\left( \beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, ,
    \label{eq:conf_probability}
\end{equation}
where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of conformation $\vec{q}$, and the integration is over the set of all possible conformations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
If the simulation is long enough, we expect the fraction of conformations $\vec{q}$ found in the bound state to converge to
\begin{equation}
    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, .
\end{equation}
After similar considerations for $P(R+L)$, we find that the ratio of visited bound and unbound conformations, in the limit of long simulations, should converge to
\begin{equation}\label{eq:bound-unbound-probability-ratio}
    \frac{P(RL)}{P(R+L)} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{unbound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} = \frac{Z(RL)}{Z(R+L)} \, ,
\end{equation}
where we have defined the \textit{configurational integral} or \textit{configurational partition function} as $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the {\textmu}s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the micosecond to second scale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations uninteresting at pharmaceutical drug discovery timescales.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig2_therm_cyc/Figure.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex, i.e. bound to a host molecule, and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent, typically water.}
    \label{fig:fig_binding_thermodynamic_cycle}
\end{figure}

@@

In practice, the mutation of $A$ to $B$ in carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimera'' molecule composed of enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Section~\ref{sec:relative-fe-protocol} for details).

@@

\paragraph{Why do alchemical calculations need unphysical intermediate states?}
While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state(see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).

To solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1})
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions is very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Section~\ref{subsec:estimators}).

@@

While absolute and relative free energy calculations have subtle differences in their practical applications (e.g., use of restraints, handling of the standard state), the fundamental ideas and concepts of relative free energy approaches remain unaltered in other types of alchemical calculations.
Absolute binding, hydration, and partition free energy still use thermodynamic cycles that enable computing transfer free energies without actually simulating the physical transfer from one environment to another.

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

>>>>

\begin{frontmatter}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Abstract
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more tractable with alchemical methods.
In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest, permits the robust computation of free energy for large transformations.
Alchemical calculations can be used in a variety of scenarios, such as: 
\begin{itemize}
\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{A});
\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 
\end{itemize}

Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
\begin{itemize}
\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; 
\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; 
\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
\end{itemize}

After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
 Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}. %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine}, and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform. 
In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
This Best Practices guide provides current recommendations and tips for users of all experience. Updates and suggestions are welcomed via our GitHub repository.

@@

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

 We assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
 Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

 While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead this guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical calculation? 
     \item How do I select an appropriate alchemical protocol? 
     \item What software tools are available to perform alchemical calculations? 
     \item How should I analyze my data and report uncertainties? 
\end{itemize}

Some other background information may be needed depending on the nature of the alchemical project. For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical simulations. This will require some basic familiarity on how to perform docking to generate reasonable simulation starting points~\cite{grinter2014challenges}.

@@

\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
$L$ and $R$ --- generic names for ligand and receptor\\
$K_b^{\circ}$ --- binding  constant \\
$c^{\circ}$ --- standard state concentration \\
$U$ --- potential energy\\
$u$ --- reduced potential describing a thermodynamic state \\
$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble)\\
$\Delta A$ --- Helmholtz free energy (canonical ensemble)\\
$\Delta f$ --- reduced (dimensionless) free energy \\
$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy\\
$\Gamma$ --- conformation space accessible by simulations \\
$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
$k_B$ --- Boltzmann constant \\
$Z$ --- partition function \\
$p$ --- pressure \\
$\mu$ --- chemical potential (grand canonical ensemble)\\
$T$ --- temperature \\
$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
$g$ --- statistical inefficiency\\
$\mathcal{O}$ --- overlap matrix\\
$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
$\tau _{eq}$ --- integrated auto-correlation time\\
$t_0$ --- equilibration time
\end{tcolorbox}

%%%%%%%%%%%%%%%%%%%%
% Theory basics    %
%%%%%%%%%%%%%%%%%%%%
\section{Statistical mechanics demonstrates why alchemical free energy calculations work}
\label{sec:theory}
Why would you want to run an alchemical free energy calculation and why do they work?
In this section, we use the example of relative free energy calculations to sketch the theory of alchemical simulations and illustrate their utility.
The emphasis here is placed on bridging theoretical foundations and intuition.
A rigorous derivation of the standard (absolute) free energy of binding using the principles of statistical mechanics can be found in Gilson's classic work~\cite{gilson1997statisticalthermodynamic}.

@@

\paragraph{The free energy of binding can be expressed as a ratio of partition functions}
A natural, though generally very computationally expensive, way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
\begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
    \Delta G_{\mathrm{bind},L} \approx \Delta A_{\mathrm{bind}, L} = - k_B T  \left( \mathrm{ln} \frac{P(RL)}{P(R+L)} + \ln \left( c^\circ N_{\mathrm{Av}} V \right) \right) \, ,
\end{equation}
where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand conformations $\vec{q}$ that we consider bound and unbound respectively.
The probability of a conformation $\vec{q}$ is given by the Boltzmann probability density function
\begin{equation}
    P(\vec{q}) = \frac{\exp\left( \beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, ,
    \label{eq:conf_probability}
\end{equation}
where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of conformation $\vec{q}$, and the integration is over the set of all possible conformations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
If the simulation is long enough, we expect the fraction of conformations $\vec{q}$ found in the bound state to converge to
\begin{equation}
    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, .
\end{equation}
After similar considerations for $P(R+L)$, we find that the ratio of visited bound and unbound conformations, in the limit of long simulations, should converge to
\begin{equation}\label{eq:bound-unbound-probability-ratio}
    \frac{P(RL)}{P(R+L)} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{unbound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} = \frac{Z(RL)}{Z(R+L)} \, ,
\end{equation}
where we have defined the \textit{configurational integral} or \textit{configurational partition function} as $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the millisecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig2_therm_cyc/Figure.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host or a rigid receptor.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex, i.e. bound to a host molecule, and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent, typically water.}
    \label{fig:fig_binding_thermodynamic_cycle}
\end{figure}

@@

In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Section~\ref{sec:relative-fe-protocol} for details).

@@

\paragraph{Why do alchemical calculations need unphysical intermediate states?}
While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state (see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).

To solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1})
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates. This is achieved by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Section~\ref{subsec:estimators}).

@@

While absolute and relative free energy calculations have subtle differences in their practical applications (e.g., use of restraints, handling of the standard state), the fundamental ideas and concepts of relative free energy approaches remain unaltered in other types of alchemical calculations.
Absolute binding, hydration, and partition free energies still use thermodynamic cycles that enable computing transfer free energies without actually simulating the physical transfer from one environment to another.

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in Section~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.
